---
title: "How Does Weather Affect Denver Bcycle Usage?"
author: "Andy Pickering"
date: "Aug 1, 2017"
output:
  html_document:
    keep_md: yes
    toc: yes
  pdf_document: default
---

## Introduction


## Bcycle Data
- <https://github.com/andypicke/Bcycle>

```{r echo=TRUE}
rm(list=ls())
#setwd("/Users/Andy/Bcycle/")
library(ggplot2)
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(dplyr))
library(ggplot2)
```


First read in the data for 2015, which I have downloaded already from <https://denver.bcycle.com/company>. Note: I tried to read in the xlsx file using the 'xlsx' package, but it didn't work. Instead I just opened excel and saved the file as a csv.

```{r ReadData}
bcyc<-read.csv("data/Bcyc2015.csv")
head(bcyc)
```


How many rides are contained in this dataset?
```{r}
nr<-nrow(bcyc)
nr
```


```{r}
str(bcyc)
```



```{r}
# add a new column of class Posixct with date/time comined
bcyc$dt_chkout<-as.POSIXct( strptime(paste(bcyc$Checkout.Date,bcyc$Checkout.Time),"%m/%d/%y %H:%M:%S"))
bcyc$dt_ret<-as.POSIXct( strptime(paste(bcyc$Return.Date,bcyc$Return.Time),"%m/%d/%y %H:%M:%S"))

bcyc$month <- month(bcyc$dt_chkout)


```


```{r}
bcyc %>% group_by(month) %>% count() %>% ggplot(aes(x=month,y=n)) +geom_point() + geom_line() + geom_smooth()
```



## Weather Data
So we can see that the total rides peaks around August, and is lowest around December. This is probably related to the weather, let's get some weather data and check this out. I'm using data downloaded from <https://www.wunderground.com>.

```{r GetWeather}

# Daily weather data for 2015
url<-"https://www.wunderground.com/history/airport/KDEN/2015/1/1/CustomHistory.html?dayend=31&monthend=12&yearend=2015&req_city=&req_state=&req_statename=&reqdb.zip=&reqdb.magic=&reqdb.wmo=&format=1"

download.file(url,"DenWeather2015.csv")
wea<-read.csv("DenWeather2015.csv")

wea$MST <- as.Date(wea$MST,"%Y-%m-%d")
wea$month <- month(wea$MS)

# in Precip "T" is trace I think; change to zero for analysis
idT<-which(wea$PrecipitationIn=="T")
wea$PrecipitationIn[idT]<-"0.00"
wea$PrecipitationIn <- as.numeric(as.character(wea$PrecipitationIn))

#wea %>% mutate(month = month(MST))

head(wea)



```


```{r}
wea %>% group_by(month) %>% summarise(Tavg = mean(`Mean.TemperatureF`)) %>% ggplot(aes(x=month,y=Tavg)) + geom_point() + geom_line() + geom_smooth()

```


## Comparison of weather and Bcycle data 
- The MaxTemp seasonal cycle looks very similar to the month ride totals. Let's make some scatterplots to better see the correlation between weather variables and the number of rides.



### Monthly


```{r}
bcyc_monthly <- bcyc %>% group_by(month) %>% count()
wea_monthly  <- wea %>% group_by(month) %>% summarise(tavg=mean(`Mean.TemperatureF`,na.rm=TRUE))
month_merge  <- merge(bcyc_monthly,wea_monthly) 
ggplot(month_merge,aes(x=tavg,y=n))+geom_point() + geom_smooth(method = "lm")
```



### Daily

```{r Compute Daily Rides}

bcyc$yday <- yday(bcyc$dt_chkout)
wea$yday <- yday(wea$MST)

bcyc %>% group_by(yday) %>%count()%>% ggplot(aes(yday,n)) + geom_point() + geom_smooth()
wea %>% ggplot(aes(yday,`Mean.TemperatureF`)) + geom_point() + geom_smooth()

```



At the daily level, the seasonal pattern is the same but there is a lot more variability, especially in the winter/spring.  
  

### Rides vs Temperature 
```{r}
bcyc_daily <- bcyc %>% group_by(yday) %>% count()
yday_merge <- merge(bcyc_daily,wea) %>% select(yday,n,`Mean.TemperatureF`)
yday_merge %>% ggplot(aes(Mean.TemperatureF,n)) + geom_point() +  geom_smooth(col='red') + geom_smooth(method = "lm")
```


### Ride duration vs temperature.

```{r Ride Durations for different temp ranges}
# Need to add the daily temp to each bcyc ride
for (i in seq_along(yday(wea$MST))){
        ig<-which(bcyc$yday == yday(wea$MST)[i])
        bcyc$temp[ig]<-wea$Max.TemperatureF[i]
}


tlim<-40
par(mfrow=c(1,3))
ig <- which(bcyc$temp<30 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="temp<30",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

ig <- which(bcyc$temp>30 & bcyc$temp<60 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="30<temp<60",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

ig <- which( bcyc$temp>60 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="temp>60",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

```


## Conclusions:  
- The total number of Denver Bcycle rides has a strong seasonal cycle, peaking around August and minimum around January.  
- The total number of Denver Bcycle rides per month is strongly correlated with the monthly mean of max temperatures.  
- Below about 30 deg and above 80 deg, the number of rides is less dependent on further decreasing(increasing) temperature.  
- The mean and median ride durations tend to be larger for increasing temperatures.  

##  Follow-up Questions:  
- Do all years look the same?  
- Does the relationship between weather and rides look different for different types of passes (ie annual vs 24 hour)?  
- Is there a stronger correlation with precip on shorter timescales (hourly?)?

