---
title: "How Does Weather Affect Denver Bcycle Usage?"
author: "Andy Pickering"
date: "May 25, 2016"
output: 
  html_document: 
    keep_md: yes
---

## <https://github.com/andypicke/Bcycle>

```{r echo=TRUE}
rm(list=ls())
setwd("/Users/Andy/Bcycle/")
library(ggplot2)
library(lubridate)
```

First read in the data for 2015, which I have downloaded already from <https://denver.bcycle.com/company>. Note: I tried to read in the xlsx file using the 'xlsx' package, but it didn't work. Instead I just opened excel and saved the file as a csv.
```{r ReadData}
bcyc<-read.csv("Bcyc2015.csv")
head(bcyc)
```

How many rides are contained in this dataset?
```{r}
nr<-nrow(bcyc)
```

So we have `r nr` observations (rides) in this dataset.


```{r}
# add a new column of class Posixct with date/time comined
bcyc$dt_chkout<-as.POSIXct( strptime(paste(bcyc$Checkout.Date,bcyc$Checkout.Time),"%m/%d/%y %H:%M:%S"))
bcyc$dt_ret<-as.POSIXct( strptime(paste(bcyc$Return.Date,bcyc$Return.Time),"%m/%d/%y %H:%M:%S"))
```


## First I want to compute the total rides per month and see what kind of seasonal cycle there is.
```{r }
# List of months
month_list <- c("January","February","March","April","May","June","July","August","September","October","November","December")
bcyc$month <- months(bcyc$dt_chkout)
tot_rides_month <- vector(mode='numeric',length=12)

for (i in seq_along(month_list)){
        a<-which(bcyc$month==month_list[i])
        tot_rides_month[i] <- length(a)
}

# Make a new dataframe w/ monthy rides
bcyc_monthly <- data.frame(rides=tot_rides_month,month=month_list,monthID=1:12)

# Plot rides per month vs. month
qplot(bcyc_monthly$monthID,bcyc_monthly$rides,xlab="Month",ylab="Total Rides",main="Total Rides by Month: 2015",geom=c("point","smooth"))
```  
  
So we can see that the total rides peaks around August, and is lowest around December. This is probably related to the weather, let's get some weather data and check this out. I'm using data downloaded from <https://www.wunderground.com>.

```{r GetWeather}
# Daily weather data for 2015
url<-"https://www.wunderground.com/history/airport/KDEN/2015/1/1/CustomHistory.html?dayend=31&monthend=12&yearend=2015&req_city=&req_state=&req_statename=&reqdb.zip=&reqdb.magic=&reqdb.wmo=&format=1"

download.file(url,"DenWeather2015.csv")

wea<-read.csv("DenWeather2015.csv")
wea$MST <- as.Date(wea$MST,"%Y-%m-%d")
wea$month <- months(wea$MST)

# in Precip "T" is trace I think; change to zero for analysis
idT<-which(wea$PrecipitationIn=="T")
wea$PrecipitationIn[idT]<-"0.00"
wea$PrecipitationIn <- as.numeric(as.character(wea$PrecipitationIn))

# compute mean weather values per month
maxtemp_mean_month <- vector(mode='numeric',length=12)
mintemp_mean_month <- vector(mode='numeric',length=12)
precip_mean_month <- vector(mode='numeric',length=12)
maxwind_mean_month <- vector(mode='numeric',length=12)

for (i in seq_along(month_list)){
        a<-which(wea$month==month_list[i])
        maxtemp_mean_month[i] <- mean(wea$Max.TemperatureF[a])
        mintemp_mean_month[i] <- mean(wea$Min.TemperatureF[a],na.rm = TRUE)
        precip_mean_month[i] <- mean(wea$PrecipitationIn[a],na.rm = TRUE)
        maxwind_mean_month[i] <- mean(wea$Max.Wind.SpeedMPH[a],na.rm = TRUE)
}

# Make a data frame w/ monthly values
W_mon=data.frame(precip=precip_mean_month,maxtemp=maxtemp_mean_month,mintemp=mintemp_mean_month,maxwind=maxwind_mean_month,monthID=1:12,month=month_list)


qplot(W_mon$monthID,W_mon$maxtemp,geom=c("point","smooth"),xlab="Month",ylab="Max Temp",main="Maximum Temps 2015")
qplot(W_mon$monthID,W_mon$precip,geom=c("point","smooth"),xlab="Month",ylab="Precip",main="Precip 2015")
qplot(W_mon$monthID,W_mon$maxwind,geom=c("point","smooth"),xlab="Month",ylab="Max Wind",main="Maximum Wind 2015")

```

## The MaxTemp seasonal cycle looks very similar to the month ride totals. Let's make some scatterplots to better see the correlation between weather variables and the number of rides.

```{r Scatter weather vs rides}
month_merge<-merge(bcyc_monthly,W_mon)
ggplot(dat=month_merge,aes(x=maxtemp,y=rides))+geom_point(shape=1)+geom_smooth(method=lm)
ggplot(dat=month_merge,aes(x=precip,y=rides))+geom_point(shape=1)+geom_smooth(method=lm)
ggplot(dat=month_merge,aes(x=maxwind,y=rides))+geom_point(shape=1)+geom_smooth(method=lm)
```


## Let's look in a little more detail at the daily level.


```{r Compute Daily Rides}

bcyc$yday <- yday(bcyc$dt_chkout)
uniq_yday <- sort(unique(bcyc$yday))

# Make an empty vector to store results in
tot_rides_daily <- vector(mode="numeric",length=length(uniq_yday))

for (i in seq_along(uniq_yday)) {
        rm(a)
        a<-which(bcyc$yday==uniq_yday[i])
        tot_rides_daily[i] <- length(a)
}

qplot(uniq_yday,tot_rides_daily,xlab="Yearday",ylab="Total Rides",main="Daily total rides",geom=c("point","smooth"))

```

At the daily level, the seasonal pattern is the same but there is a lot more variability, especially in the winter/spring.  
  
  Let's look at the relationship between daily temperature and rides.

```{r}
qplot(wea$Max.TemperatureF,tot_rides_daily,xlab="Max Temp",ylab="Total Rides",main="Daily total rides vs Temp",geom=c("point","smooth"))

```

## Look at how ride durations change w/ temperature.

```{r Ride Durations for different temp ranges}
# Need to add the daily temp to each bcyc ride
for (i in seq_along(yday(wea$MST))){
        ig<-which(bcyc$yday == yday(wea$MST)[i])
        bcyc$temp[ig]<-wea$Max.TemperatureF[i]
}


tlim<-40
par(mfrow=c(1,3))
ig <- which(bcyc$temp<30 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="temp<30",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

ig <- which(bcyc$temp>30 & bcyc$temp<60 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="30<temp<60",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

ig <- which( bcyc$temp>60 & bcyc$Duration..Minutes.<tlim)
hist(bcyc$Duration..Minutes.[ig],main="temp>60",xlab="Duration (min)")
rug(bcyc$Duration..Minutes.[ig])
abline(v=median(bcyc$Duration..Minutes.[ig]),col="red")
abline(v=mean(bcyc$Duration..Minutes.[ig]),col="blue")

```


#Conclusions:  
- The total number of Denver Bcycle rides has a strong seasonal cycle, peaking around August and minimum around January.  
- The total number of Denver Bcycle rides per month is strongly correlated with the monthly mean of max temperatures.  
- Below about 30 deg and above 80 deg, the number of rides is less dependent on further decreasing(increasing) temperature.  
- The mean and median ride durations tend to be larger for increasing temperatures.  

# Follow-up Questions:  
- Do all years look the same?  
- Does the relationship between weather and rides look different for different types of passes (ie annual vs 24 hour)?  
- Is there a stronger correlation with precip on shorter timescales (hourly?)?

